Urban and socio-economic correlates of property prices in Dublin's area

### 2020-08-18 manuscript/manuscript.Rmd
reviewer changes

---
title: Urban and socio-economic correlates of property price evolution\:\ Application to Dublin's costal area
affiliation:
  institution-columnar: true

@@

keywords: ["Property price", "Housing Market", "Feature Analysis", "Machine Learning", "Geocoding"]
abstract: |
  Understanding the characteristics of the housing market is essential for both sellers and buyers. However, the housing market is influenced by multiple factors. In this paper, the urban and socio-economic structure of an area is used to predict the price of 10389 houses sold in 2018 in the city of Dublin. More precisely, the direct distance from each house to 160 urban features taken from OpenStreetMap is calculated, and an extreme gradient boosting linear regression performed. Using these features, the model explains 47\% of the housing price variance. The most important features in this model are the proximity to an embassy and to a grassland. In addition, the results of a population census from 2016 are also used to correlate with the price of houses. From this census, 48 features are used as the input of a gradient boost linear regression model. In all, the socio-economic features are explaining 42\% of the housing price variance as well. The most important socio-economic features are; the density of houses having more than eight or more rooms in the area, the density of young children, and the density of individuals reporting that they have no religion. By taking into account either urban or socio-economic features, it is possible to accurately estimate housing prices and to predict their evolution.

@@

# output: word_document

header-includes:
  - \usepackage{graphicx}
---

(ref:distrib-plot) Overall (A) and geographic (B) distribution of the houses prices density in the city of Dublin in 2018. The geographic distribution was obtained by using a generalized additive model with soap film smooth parameter to take into account the influence of coastal boundaries. Geographic estimates outside of the 95\% Confidence Interval were removed.
(ref:osm-features-xgb) Property price prediction accuracy (A) and Property price prediction error (B) using urban features with XGBoost.
(ref:osm-features-table) Urban features importance (higher than 1\%).
(ref:census-features-xgb) Property price prediction accuracy (A) and Property price prediction error (B) using socio-economic features with XGBoost.
(ref:census-features-table) Socio-economic features importance (higher than 1\%).

```{r setup, include = FALSE}
# options ----------------------------------------------------------------------
set.seed(123)
options(scipen=999)
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE
  )

@@

# functions --------------------------------------------------------------------
source(here::here("script/functions_project.R"))
source(here::here("script/osm_dublin_features.R"))

@@

# all price registry------------------------------------------------------------
data_dublin_geocoded <- file.path(data_path, "data_dublin_geocoded_clean.rds") %>% 
  readr::read_rds() %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    year_fact = as.factor(as.character(year))
  )

@@

# dublin map -------------------------------------------------------------------
# boundaries_dublin <-
#   osmdata::opq(bbox = 'Dublin, Ireland') %>%
#   osmdata::add_osm_feature(key = 'admin_level', value = '7') %>%
#   osmdata::osmdata_sf() %>%
#   osmdata::unique_osmdata()
# 
# bb <- st_bbox(boundaries_dublin$osm_multipolygons$geometry)
# # Defining a buffer
# buffer <- 0.1
# p_big <- rbind(
#   c(bb[1] - buffer, bb[2] - buffer),
#   c(bb[1] - buffer, bb[4] + buffer),
#   c(bb[3] + buffer, bb[4] + buffer),
#   c(bb[3] + buffer, bb[2] - buffer),
#   c(bb[1] - buffer, bb[2] - buffer)
#   )
# # Putting the coordinates into a squared polygon object
# pol_dublin <- st_polygon(list(p_big)) %>% st_geometry
# # Providing the SRID (here, unprojected lon/lat)
# st_crs(pol_dublin) <- 4326
# 
# # get Dublin coast line ------------------------------------------------------
# coast_dublin <- opq(bbox = st_bbox(pol_dublin)) %>%
#   add_osm_feature(key = 'natural', value = 'coastline') %>%
#   osmdata_sf %>% unique_osmdata
# 
# blade_dublin <- coast_dublin$osm_lines$geometry %>% st_union %>% st_line_merge
# 
# ls <- st_linestring(blade_dublin[[1]][[1]])
# 
# dublin_multipol <- st_split(st_geometry(pol_dublin), st_geometry(ls))
# 
# readr::write_rds(dublin_multipol, here::here("data/dublin_multipol.rds"))
dublin_multipol <- readr::read_rds(here::here("data/dublin_multipol.rds"))

@@

Beside their inherent characteristics such as size, design and materials [@bourassa2007spatial; @liu2013spatial], property valuation is dictated by multiple external factors explaining their spatial autocorrelation [@basu1998analysis]. Some of these factors are macroeconomic [@antonakakis2016dynamic], related to the evolution of population general wealth while others are related to the characteristics of the neighbourhood [@dubin1998predicting]. These characteristics include the urban and socio economic structure of the neighbourhood [@goodman1998housing]. However, it is difficult to evaluate the influence of such external factors on property valuation [@clapp2002predicting]. By using machine learning algorithms, prediction of house prices has become more and more accurate. Techniques used usually include artificial neural networks due to the volatility of the market [@limsombunchai2004house; @feng2015comparing; @selim2009determinants].

In this paper, an alternative machine learning algorithm is used to identify urban and socio-economic correlates of property price evolution in the area of Dublin, Ireland in 2018. Due to the high number of potential features, an eXtreme Gradient Boosting (XGBoost) regression model is believed to obtain the best results with low parameter customisation and high speed.

@@

Because of the evolution of urban planning and the evolution of socio-economic features measured by the 2016 census, only houses sold in 2018 were processed to avoid potential temporal incoherences. In 2018, a total of 10395 property in the area of Dublin were sold, and these constitute the database used to identify the urban and socio-economic correlates of property prices.

In order to evaluate the spacial distribution of the property sold, property addresses were geocoded (i.e., converted to latitude and longitude) using the OpenStreetMap API. OpenStreetMap is a collaborative project which aims to create and provide access to free editable maps of the world [@haklay2008openstreetmap].

To estimate the geographical distribution of the price density, multiple methods such as multiple regression [@mccluskey2000application] or Bayesian smoothing [@clapp2002predicting] have been employed. While these methods are efficient in a non-restricted space, they have limitations when they are used in coastal areas. To deal with the influence of boundaries on price estimation, a generalised additive model using soap film smoothing has been used [@wood2008soap].

@@

OpenStreetMap combines information about more than 400 urban features including road information and building information to categorize features such as amenities, leisure or tourism structures [@openstreetmap2020]. Among these features, only 160 are available or relevant to the Dublin Area. The distance between each property and the closest point corresponds to each of the 160 urban features.

@@

In addition to the distance to every urban feature, each house sold in 2018 was related to 48 socio-economic features of the small area including the house, as measured by the Irish census 2016 [@cso2020]. The results of Irish 2016 census consultation is accessible through both the Central Statistics Office and All-Island Research Observatory. The data obtained can be mapped over small area boundaries which are fractions of Irish Electoral Division map. The social features extracted are corresponding to population information, religion, carers and health. Economic features correspond to the characteristics of each small area; including the proportion of housing types, number of rooms, occupancy and tenure per small area. Each property is then associated to the value corresponding to its small area. For anonymisation purposes, the results from small areas having less than five respondents to the census were converted to a proportion of five respondents.

## Gradient Boosting Regressions

Lastly, an eXtreme Gradient Boosting (XGBoost) regression model was calculated to examine the extent to which house prices rely on urban and on socio-economic features. The XGBoost algorithm optimizes the prediction accuracy by performing iterative least-squares regressions (set to 100 iterations as the best trade-off between accuracy and speed), thereby minimizing the root mean squared error [@friedman2001greedy; @chen2016xgboost]. The original dataset has been randomly split to 80% for training and 20% for testing the models. Models accuracy is estimated using the coefficient of determination ($R^2$), the Root Mean Square Error ($RMSE$) and the Mean Absolute Error ($MAE$) of the predicted property price values of the test dataset.

@@

In 2018, the average price of a sold property in Dublin was `r scales::dollar(round(mean_dat, 0), suffix = "", prefix = "€")` (SD = `r scales::dollar(round(sd_dat, 0), suffix = "", prefix = "€")`). In order to remove potential human errors and outliers, prices higher or lower than 1 SD were removed from the original dataset.

```{r distrib-plot, fig.env='figure*', fig.height=3, out.width='0.98\\textwidth', fig.cap="(ref:distrib-plot)"}
price_density <- data_dublin_geocoded %>%
  dplyr::filter(year == 2018) %>%
  magrittr::use_series(price) %>%
  stats::density() %$%
  data.frame(x, y) %>%
  ggplot() +
  geom_segment(aes(
    x = x,
    xend = x,
    y = 0,
    yend = y,
    color = x
  )) +
  scale_x_continuous("Price", labels = scales::dollar_format(suffix = "", prefix = "€")) +
  scale_y_continuous("Density") +
  scale_color_gradient(
    low = '#3a53a4', 
    high = '#ed1c24', 
    ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 6),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  ) +
  ggtitle("A")

map_density <- gam_pred %>%
  dplyr::filter(value > 200000) %>%
  ggplot() +
  geom_raster(aes(x = lng, y = lat, fill = value)) +
  geom_sf(data = dublin_sea, fill = "light blue") +
  scale_fill_gradient(
    "Price Density", 
    low = '#3a53a4', 
    high = '#ed1c24', 
    trans = "log",
    breaks = c(250000, 1000000), 
    labels = scales::dollar_format(suffix = "", prefix = "€")) +
  scale_x_continuous(
    "Longitude", 
    limits = c(-6.5, -6), 
    breaks = seq(-6.5, -6, 0.2)) +
  scale_y_continuous(
    "Latitude", 
    limits = c(53.2, 53.5), 
    breaks = seq(53.2, 53.5, 0.1)) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 6),
    legend.position = "none"
  ) +
  ggtitle("B")

price_density + map_density
```

The density of housing prices reveals a unimodal distribution of property prices (Figure \@ref(fig:distrib-plot)A). In addition, the geographical distribution of property prices indicates a higher valuation in the South-West of the city as well as on the coast line (Figure \@ref(fig:distrib-plot)B).

@@

test_tbl_y <- test_dublin_2018_features %>%
  dplyr::select(price) %>%
  as.matrix()
test_tbl_x <- test_dublin_2018_features %>%
  dplyr::select(any_of(osm_dublin_features)) %>% 
  as.matrix()

xgmodel <- xgboost::xgboost(
  data = train_tbl_x,
  label = train_tbl_y,
  objective = "reg:squarederror",
  eval_metric = "rmse",
  verbose = 0,
  nround = 100
)

@@

Using the XGBoost algorithm, the 160 urban features are explaining `r scales::percent(metrics_result$.estimate[metrics_result$.metric == "rsq"])` of the property price variance (`r lm_result$statistic$modelfit$r2`) with a $RMSE$ of `r scales::dollar(metrics_result$.estimate[metrics_result$.metric == "rmse"], suffix = "", prefix = "€")` and a $MAE$ of `r scales::dollar(metrics_result$.estimate[metrics_result$.metric == "mae"], suffix = "", prefix = "€")`.

```{r osm-features-xgb, fig.cap="(ref:osm-features-xgb)", fig.height=8, out.width='0.98\\columnwidth'}
result_accuracy <- result %>%
  ggplot(aes(x = price, y = pred)) +
  geom_point(alpha = 0.1) +
  geom_smooth(
    method = "lm",
    fullrange = TRUE,
    se = TRUE,
    color = "black",
    size = 0.1
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif"),
    legend.position = "none"
  ) +
  labs(
    title = "A",
    subtitle = latex2exp::TeX(paste(
      "$R^2 =$", 
      scales::percent(metrics_result$.estimate[metrics_result$.metric == "rsq"]),
      ";",
      "$RMSE =$",
      scales::dollar(
        metrics_result$.estimate[metrics_result$.metric == "rmse"], 
        suffix = "", 
        prefix = "€"
      ),
      ";",
      "$MAE =$",
      scales::dollar(
        metrics_result$.estimate[metrics_result$.metric == "mae"], 
        suffix = "", 
        prefix = "€"
      )
      )
    )
  )

result_error <- result %>%
  dplyr::mutate(diff_percent = (pred - price) / price) %>%
  ggplot(aes(x = price, y = diff_percent)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(
    color = "black",
    se = TRUE,
    fullrange = TRUE,
    size = 0.1
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price Error",
    limits = c(-0.5, 0.5),
    labels = scales::percent
  ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif")
  ) +
  ggtitle("B")

@@

The overall correlation of the predicted property prices with urban features is shown in Figure \@ref(fig:osm-features-xgb)A. It can be noticed that property prices situated at the low and the high end of the range are the most difficult to predict (Figure \@ref(fig:osm-features-xgb)B). A possible explanation is the absence of distinctive and recurrent patterns in urban features for these houses. However, the prices higher than €300,000 are potentially driving down the prediction accuracy due to outliers.

```{r osm-features-table, results="asis"}
table_osm <- xgb.importance(model = xgmodel) %>%
  dplyr::filter(Gain > 0.01) %>%
  dplyr::arrange(desc(Gain)) %>%
  dplyr::mutate(across(where(is.numeric), scales::percent, accuracy = 0.1)) %>%
  dplyr::select(Feature, Importance = Gain) %>%
  tidyr::separate(Feature, c("Feature Category", "Feature Type"), sep = "_")

table_osm %>% 
  knitr::kable(
    "latex",
    caption = "(ref:osm-features-table)",
    booktabs = TRUE,
    linesep = ""
  ) %>%
  kableExtra::kable_styling(font_size = 8, full_width = TRUE)
```

@@

@@

```{r census-features}
# census dublin features
source(here::here("script/census_dublin_features.R"))

train_tbl_y <- train_dublin_2018_features %>%
  dplyr::select(price) %>%
  as.matrix()
train_tbl_x <- train_dublin_2018_features %>%
  dplyr::select(any_of(census_dublin_features)) %>% 
  as.matrix()

test_tbl_y <- test_dublin_2018_features %>%
  dplyr::select(price) %>%
  as.matrix()
test_tbl_x <- test_dublin_2018_features %>%
  dplyr::select(any_of(census_dublin_features)) %>% 
  as.matrix()

xgmodel <- xgboost::xgboost(
  data = train_tbl_x,
  label = train_tbl_y,
  objective = "reg:squarederror",
  eval_metric = "rmse",
  verbose = 0,
  nround = 100
)

>>>>

---
title: Urban and socio-economic correlates of property prices in Dublin's area
affiliation:
  institution-columnar: true

@@

keywords: ["Property price", "Housing Market", "Feature Analysis", "Machine Learning", "Geocoding"]
abstract: |
  Understanding the characteristics of the housing market is essential for both sellers and buyers. However, the housing market is influenced by multiple factors. In this paper, the urban and socio-economic structure of an area is used to predict the price of 10387 properties sold in 2018 in the city of Dublin. More precisely, the direct distance from each property to 160 urban features taken from OpenStreetMap is calculated, and an extreme gradient boosting linear regression performed. Using these features, the model explains 45\% of the housing price variance. The most important features in this model are the proximity to an embassy and to a grassland. In addition, the results of a population census from 2016 are also used to correlate with the price of properties. From this census, 48 features are used as the input of a gradient boost linear regression model. In all, the socio-economic features are explaining 43\% of the housing price variance as well. The density of individuals reporting that they are not providing unpaid personal help for a friend or family member as well as individuals reporting that they have no religion are the most important socio-economic feratures. By taking into account either urban or socio-economic features, it is possible to accurately estimate housing prices and to predict their evolution.

@@

header-includes:
  - \usepackage{graphicx}
  - \usepackage{flushend}
---

(ref:distrib-plot) Overall distribution of the property prices density in the city of Dublin in 2018
(ref:distrib-gps) Geographic distribution of the property prices density in the city of Dublin in 2018 using a generalized additive model with soap film smooth parameter to take into account the influence of coastal boundaries. Estimates outside of the 95\% Confidence Interval were removed.
(ref:osm-model-comparison) Accuracy of XGBoost models using urban features according the type of gradient boosting method and output objective.
(ref:osm-features-xgb) Property price prediction accuracy (A) and prediction error (B) using urban features with XGBoost.
(ref:osm-features-table) Top 10 most important urban features contributing to the property price prediction.
(ref:census-model-comparison) Accuracy of XGBoost models using socio-economic features according the type of gradient boosting method and output objective.
(ref:census-features-xgb) Property price prediction accuracy (A) and prediction error (B) using socio-economic features with XGBoost.
(ref:census-features-table) Top 10 most important socio-economic features contributing to the property price prediction.

```{r setup, include = FALSE}
# options ----------------------------------------------------------------------
base::set.seed(123)
base::options(
  scipen = 999,
  knitr.kable.NA = " - "
  )
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE,
  cache = TRUE,
  fig.retina = 3
  )

@@

# sources ----------------------------------------------------------------------
source(here::here("script/functions_project.R"))
source(here::here("script/osm_dublin_features.R"))
source(here::here("script/census_dublin_features.R"))

# data -------------------------------------------------------------------------
geocode_dublin_2018 <- here::here("data/ppr_dublin_2018.csv") %>% 
  readr::read_csv()

house_dublin_2018 <- geocode_dublin_2018 %>% 
  dplyr::filter(!grepl('apartment|Apartment|APART|APT|Apt', Address))

final_dublin_2018 <- here::here("data/data_dublin_2018_osm_features.rds") %>% 
  readr::read_rds()

@@

# all price registry------------------------------------------------------------
data_dublin_geocoded <- file.path(data_path, "data_dublin_geocoded_clean.rds") %>% 
  readr::read_rds() %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    year_fact = as.factor(as.character(year))
  ) %>% 
  dplyr::filter(year == 2018)

@@

# dublin map -------------------------------------------------------------------
dublin_multipol <- readr::read_rds(here::here("data/dublin_multipol.rds"))

@@

Beside their inherent characteristics such as size, design and materials [@bourassa2007spatial; @liu2013spatial], property valuation is dictated by multiple external factors explaining their spatial autocorrelation [@basu1998analysis]. Some of these factors are macroeconomic [@antonakakis2016dynamic], related to the evolution of population general wealth while others are related to the characteristics of the neighbourhood [@dubin1998predicting]. These characteristics include the urban and socio economic structure of the neighbourhood [@goodman1998housing]. However, it is difficult to evaluate the influence of such external factors on property valuation [@clapp2002predicting]. By using machine learning algorithms, prediction of property prices has become more and more accurate. Techniques used usually include artificial neural networks due to the volatility of the market [@limsombunchai2004house; @feng2015comparing; @selim2009determinants]. Indeed, artificial neural network techniques can quickly adjust to volatility changes, therefore incorporating sharp increases or decreases in their predictions [@aragones2007neural]. However, the model obtained from most artificial neural networks are hardly interpretable [@rudin2019stop].

In this paper, an alternative machine learning algorithm is used to identify urban and socio-economic correlates of property price evolution in the area of Dublin, Ireland in 2018. Due to the high number of potential features, an eXtreme Gradient Boosting (XGBoost) regression model is believed to obtain the best results with low parameter customisation and high speed [@chen2016xgboost]. Contrary to most artificial neural network techniques, XGBoost is an interpretation-focused method which automatically provides estimates of feature importance from a trained predictive model.

@@

Because of the evolution of urban planning and the evolution of socio-economic features measured by the 2016 census, only properties sold in 2018 were processed to avoid potential temporal incoherence. The total amount of property sold in 2018 in the county Dublin which appears in the Property Price Register is `r nrow(geocode_dublin_2018)`. From these, all property corresponding to apartments were removed (`r scales::percent(1-(nrow(house_dublin_2018)/nrow(geocode_dublin_2018)), accuracy = 0.1)` of the original dataset). Finally, unsuccessfully geocoded properties were also removed from the original dataset (`r scales::percent(1-(nrow(final_dublin_2018)/nrow(house_dublin_2018)), accuracy = 0.1)`) to obtain the `r nrow(final_dublin_2018)` of the final dataset used to identify the urban and socio-economic correlates of property prices.

In order to evaluate the spacial distribution of the property sold, property addresses were geocoded (i.e., converted to latitude and longitude) using the Nominatim API [@nominatim2020] to OpenStreetMap data. OpenStreetMap is a collaborative project which aims to create and provide access to free editable maps of the world [@haklay2008openstreetmap].

To estimate the geographical distribution of the price density, multiple methods such as multiple regression [@mccluskey2000application] or Bayesian smoothing [@clapp2002predicting] have been employed. While these methods are efficient in a non-restricted space, they have limitations when they are used in coastal areas. This problem is known as "finite area smoothing" and occurs when predictions from a model are approximated across geographical barriers, such as irregular coastal shapes, which would lead to poorly predicted values [@ramsay2002spline]. To deal with the influence of geographical boundaries on price estimation, a generalised additive model using soap film smoothing has been used [@wood2008soap].

@@

OpenStreetMap combines information about more than 400 urban features including road information and building information to categorize features such as amenities, leisure or tourism structures [@openstreetmap2020]. Among these features, only 160 are available or relevant to the Dublin Area. Urban features are obtained using the R package osmdata [@R-osmdata] which queries the Overpass API [@wiki2010overpass]. The distance between each property and the closest point corresponds to each of the 160 urban features.

@@

In addition to the distance to every urban feature, each property sold in 2018 was related to 48 socio-economic features of the small area including the property, as measured by the Irish census 2016. The results of Irish 2016 census consultation is accessible through both the Central Statistics Office [@cso2020] and All-Island Research Observatory [@airo2020]. The data obtained can be mapped over small area boundaries which are fractions of Irish Electoral Division map. The social features extracted are corresponding to population information, religion, carers and health. Economic features correspond to the characteristics of each small area; including the proportion of housing types, number of rooms, occupancy and tenure per small area. Each property is then associated to the value corresponding to its small area. For anonymisation purposes, the results from small areas having less than five respondents to the census were converted to a proportion of five respondents.

## eXtreme Gradient Boosting

Lastly, an eXtreme Gradient Boosting (XGBoost) model was calculated to examine the extent to which property prices rely on urban and on socio-economic features. The XGBoost algorithm optimizes the prediction accuracy by performing boosting iterations (set to 100 iterations as the best trade-off between accuracy and speed), thereby minimizing the root mean squared error when the model's objective is a continuous output [@friedman2001greedy]. The R package "xgboost" (v.1.1.1.1) is used to implement the XGBoost algorithm [@chen2016xgboost]. With this version, two types of booster are available: tree based and linear regression based. Different types of learning task can also be defined according to the distribution of the output. In the case of a continuous output, they include regression with squared loss, regression with squared log loss, gamma regression with log-link, and tweedie regression with log-link.

The original dataset has been randomly split to 80% for training and 20% for testing the models. The prediction accuracy is estimated using the coefficient of determination ($R^2$), the Root Mean Square Error ($RMSE$) and the Mean Absolute Error ($MAE$) of the predicted property price values of the test dataset.

Both tree and linear regression based boosting model are used for every of the continuous output learning task in order to compare their accuracy with urban and socio-economic features. As no specific weight or bias in the models has been hypothesized, the parameters of the XGBoost function have been set to the default value. Then, the most accurate model is used to identify the features that are contributing the most to the prediction of property prices. This importance is a measure of the improvement in accuracy brought by a feature based on its coefficient's absolute magnitude.

@@

In 2018, the average price of a sold property in Dublin was `r scales::dollar(round(mean_dat, 0), suffix = "", prefix = "€")` with a standard deviation of `r scales::dollar(round(sd_dat, 0), suffix = "", prefix = "€")`. In order to remove potential human errors and outliers, prices higher or lower than 1 SD were removed from the original dataset.

```{r distrib-plot, fig.height=3, out.width="0.98\\columnwidth", fig.pos ="!h", fig.cap="(ref:distrib-plot)"}
data_dublin_geocoded %>%
  dplyr::filter(year == 2018) %>%
  magrittr::use_series(price) %>%
  stats::density() %$%
  data.frame(x, y) %>%
  ggplot() +
  geom_segment(aes(
    x = x,
    xend = x,
    y = 0,
    yend = y,
    color = x
  )) +
  scale_x_continuous("Price", labels = scales::dollar_format(suffix = "", prefix = "€")) +
  scale_y_continuous("Density") +
  scale_color_gradient(
    low = '#3a53a4', 
    high = '#ed1c24', 
    ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 12),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  )
```

```{r distrib-gps, fig.height=5, out.width="0.98\\columnwidth", fig.pos ="!h", fig.cap="(ref:distrib-gps)"}
gam_pred %>%
  dplyr::filter(value > 200000) %>%
  ggplot() +
  geom_raster(aes(x = lng, y = lat, fill = value)) +
  geom_sf(data = dublin_sea, fill = "light blue") +
  scale_fill_gradient(
    "Price Density", 
    low = '#3a53a4', 
    high = '#ed1c24', 
    trans = "log",
    breaks = c(250000, 1000000), 
    labels = scales::dollar_format(suffix = "", prefix = "€")) +
  scale_x_continuous(
    "Longitude", 
    limits = c(-6.5, -6), 
    breaks = seq(-6.5, -6, 0.2)) +
  scale_y_continuous(
    "Latitude", 
    limits = c(53.2, 53.5), 
    breaks = seq(53.2, 53.5, 0.1)) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 12),
    legend.position = "none"
  )
```

The density of housing prices reveals a unimodal distribution of property prices (Figure \@ref(fig:distrib-plot)). In addition, the geographical distribution of property prices indicates a higher valuation in the South-West of the city as well as on the coast line (Figure \@ref(fig:distrib-gps)).

@@

test_tbl_y <- test_dublin_2018_features %>%
  dplyr::select(price) %>%
  as.matrix()
test_tbl_x <- test_dublin_2018_features %>%
  dplyr::select(any_of(osm_dublin_features)) %>% 
  as.matrix()
```

In order to compare the possibilities of the different type of parameters in the XGBoost models,
all combinations of tree and linear regression based boosters with all possible output objectives (i.e squared error, squared log error, gamma, and tweedie) are displayed Table \@ref(tab:osm-model-comparison).

```{r osm-model-comparison, results="asis"}
comparison_xgboost <- 
  tidyr::expand_grid(
    booster_parameter = c("gbtree", "gblinear"), 
    objective_parameter = c("reg:squarederror", "reg:squaredlogerror", "reg:gamma", "reg:tweedie")
  ) %>%
  dplyr::rowwise() %>%
  dplyr::summarise(xgboost_all_parameters(booster_parameter, objective_parameter)) %>% 
  tidyr::pivot_wider(names_from = .metric, values_from = .estimate) %>% 
  dplyr::mutate(
    objective = stringr::str_sub(objective, 5),
    rmse = scales::dollar(rmse, suffix = "", prefix = "€"),
    mae = scales::dollar(mae, suffix = "", prefix = "€")
  )

comparison_xgboost %>% 
  dplyr::select(-booster) %>% 
  knitr::kable(
    "latex",
    caption = "(ref:osm-model-comparison)",
    booktabs = TRUE,
    linesep = "",
    col.names = c("Objective", "$RMSE$", "$R^2$", "$MAE$"),
    escape = FALSE
  ) %>%
  kableExtra::kable_styling(font_size = 8, latex_options = "HOLD_position") %>% 
  kableExtra::pack_rows(index = c("Tree based" = 4, "Linear regression based" = 4)) %>% 
  kableExtra::footnote("Missing value ares due to incompatibilities between data and methods.")
```

The model comparison reveals that tree based boosting models are more accurate than linear regression based model. In addition, within the possible objectives of the tree based boosting models, the tweedie regression with log-link appears to obtain the highest $R^2$ value as well as the lowest $RMSE$ and $MAE$ prediction errors.

```{r osm-best-model}
xgmodel <- xgboost::xgboost(
  # https://xgboost.readthedocs.io/en/latest/parameter.html --------------------
  data = train_tbl_x,
  label = train_tbl_y,
  # Parameters for Tree Booster ------------------------------------------------
  booster = "gbtree",
  learning_rate = 0.3,
  min_split_loss = 0,
  max_depth = 6,
  min_child_weight = 1,
  max_delta_step = 0,
  subsample = 1,
  sampling_method = "uniform",
  colsample_bytree = 1,
  colsample_bylevel = 1,
  colsample_bynode = 1,
  reg_lambda = 1,
  reg_alpha = 0,
  tree_method = "auto", # Choices: auto, exact, approx, hist, gpu_hist
  # scale_pos_weight = 1,
  refresh_leaf = 1,
  process_type = "default", # Choices: default, update
  grow_policy = "depthwise", # Choices: depthwise, lossguide
  max_leaves = 0,
  # Parameters for Linear Booster ----------------------------------------------
  # booster = "gblinear",
  # reg_lambda = 0,
  # reg_alpha = 0,
  # updater = "shotgun", # Choices: shotgun, coord_descent
  # feature_selector = "cyclic", # Choices: cyclic, shuffle
  # Learning Task Parameters ---------------------------------------------------
  objective = "reg:tweedie", 
  # Choices (reg only): squarederror, squaredlogerror, gamma, tweedie
  eval_metric = "rmse",
  # Choices: rmse, rmsle, mae, mphe, logloss, error, error@t, merror, ...
  base_score = 0.5,
  verbose = 0,
  nround = 100
)

@@

Using the XGBoost algorithm with a tree based booster and a tweedie regression with log-link learning objective, the 160 urban features are explaining `r scales::percent(metrics_result$.estimate[metrics_result$.metric == "rsq"])` of the property price variance (`r lm_result$statistic$modelfit$r2`) with a $RMSE$ of `r scales::dollar(metrics_result$.estimate[metrics_result$.metric == "rmse"], suffix = "", prefix = "€")` and a $MAE$ of `r scales::dollar(metrics_result$.estimate[metrics_result$.metric == "mae"], suffix = "", prefix = "€")`.

The overall correlation of the predicted property prices with urban features is shown in Figure \@ref(fig:osm-features-xgb)A. It can be noticed that property prices situated at the low and the high end of the range are the most difficult to predict (Figure \@ref(fig:osm-features-xgb)B). A possible explanation is the absence of distinctive and recurrent patterns in urban features for these properties. However, the prices higher than €300,000 are potentially driving down the prediction accuracy due to outliers.

```{r osm-features-xgb, fig.cap="(ref:osm-features-xgb)", fig.height=7, fig.pos ="!h", out.width="0.98\\columnwidth"}
result_accuracy <- result %>%
  ggplot(aes(x = price, y = pred)) +
  geom_point(alpha = 0.1) +
  geom_smooth(
    method = "lm",
    fullrange = TRUE,
    se = TRUE,
    color = "black",
    size = 0.1
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 12),
    legend.position = "none"
  ) +
  labs(
    title = "A",
    subtitle = latex2exp::TeX(paste(
      "$R^2 =$", 
      scales::percent(metrics_result$.estimate[metrics_result$.metric == "rsq"]),
      ";",
      "$RMSE =$",
      scales::dollar(
        metrics_result$.estimate[metrics_result$.metric == "rmse"], 
        suffix = "", 
        prefix = "€"
      ),
      ";",
      "$MAE =$",
      scales::dollar(
        metrics_result$.estimate[metrics_result$.metric == "mae"], 
        suffix = "", 
        prefix = "€"
      )
      )
    )
  )

result_error <- result %>%
  dplyr::mutate(diff_percent = (pred - price) / price) %>%
  ggplot(aes(x = price, y = diff_percent)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(
    color = "black",
    se = TRUE,
    fullrange = TRUE,
    size = 0.1
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price Error",
    limits = c(-0.5, 0.5),
    labels = scales::percent
  ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 12)
  ) +
  ggtitle("B")

@@

```{r features-table}
table_osm <- xgboost::xgb.importance(model = xgmodel) %>%
  dplyr::slice_head(n = 10) %>% 
  #dplyr::filter(Gain > 0.01) %>%
  dplyr::arrange(desc(Gain)) %>%
  dplyr::mutate(across(where(is.numeric), scales::percent, accuracy = 0.1)) %>%
  dplyr::select(Feature, Importance = Gain) %>%
  tidyr::separate(Feature, c("Feature Category", "Feature Type"), sep = "_")
```

@@

```{r osm-features-table, results="asis"}
table_osm %>% 
  knitr::kable(
    "latex",
    caption = "(ref:osm-features-table)",
    booktabs = TRUE,
    linesep = ""
  ) %>%
  kableExtra::kable_styling(
    font_size = 8, 
    full_width = TRUE, 
    latex_options = "HOLD_position"
  )
```

@@

```{r census-features}
train_tbl_y <- train_dublin_2018_features %>%
  dplyr::select(price) %>%
  as.matrix()
train_tbl_x <- train_dublin_2018_features %>%
  dplyr::select(any_of(census_dublin_features)) %>% 
  as.matrix()

test_tbl_y <- test_dublin_2018_features %>%
  dplyr::select(price) %>%
  as.matrix()
test_tbl_x <- test_dublin_2018_features %>%
  dplyr::select(any_of(census_dublin_features)) %>% 
  as.matrix()
```

Similarly to urban features, the accuracy of different XGBoost models using socio-economic features to predict property prices are displayed Table \@ref(tab:census-model-comparison).

```{r census-model-comparison, results="asis"}
comparison_xgboost <- 
  tidyr::expand_grid(
    booster_parameter = c("gbtree", "gblinear"), 
    objective_parameter = c("reg:squarederror", "reg:squaredlogerror", "reg:gamma", "reg:tweedie")
  ) %>%
  dplyr::rowwise() %>%
  dplyr::summarise(xgboost_all_parameters(booster_parameter, objective_parameter)) %>% 
  tidyr::pivot_wider(names_from = .metric, values_from = .estimate) %>% 
  dplyr::mutate(
    objective = stringr::str_sub(objective, 5),
    rmse = scales::dollar(rmse, suffix = "", prefix = "€"),
    mae = scales::dollar(mae, suffix = "", prefix = "€")
  )

comparison_xgboost %>% 
  dplyr::select(-booster) %>% 
  knitr::kable(
    "latex",
    caption = "(ref:census-model-comparison)",
    booktabs = T,
    linesep = "",
    col.names = c("Objective", "$RMSE$", "$R^2$", "$MAE$"),
    escape = FALSE
  ) %>%
  kableExtra::kable_styling(font_size = 8, latex_options = "HOLD_position") %>% 
  pack_rows(index = c("Tree based" = 4, "Linear regression based" = 4)) %>% 
  kableExtra::footnote("Missing value ares due to incompatibilities between data and methods.")
```
